---
title: "Ordinating categorical data using Vegan"
author: "Jari Oksanen, adapted by T. J. Smith"
date: "`r Sys.Date()`"
output: html_document
---

## Summary

This vignette describes some of the most commonly used ordination pathways supported by this package. Unconstrained ordination examples cover principal coordinate analysis, detrended correspondence analysis, and non-metric multidimensional scaling. This vignette shows how to interpret their results by fitting environmental vectors and factors or smooth environmental surfaces to the graph. The basic plotting command, and more advanced plotting commands for congested plots are also discussed, as well as adding items such as ellipses, convex hulls, and other items for classes. The constrained ordination example employs constrained (canonical) correspondence analysis. It shows how a model is defined, and discusses model building and significance tests of the whole ordination, single constraints, and axes.

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
